---
output:
  pdf_document: default
  html_document: default
---

# Sampling objects in a dataset

In this activity, we will demonstrate two methods of sampling objects in a dataset:

- Naive random sampling
- Stratified random sampling

Sampling is a common method of dealing with a huge number of objects/observations in a dataset.

## Dependencies and setup

In this activity, we'll be using the following R packages:

```{r}
library(tidyverse)
```

We'll sample from our handy-dandy Star Wars data set:

```{r}
head(starwars)
```

## Random sampling

In naive random sampling, we simply randomly sample rows from our dataset.
In this example, we'll sample without replacement: we don't want to characters more than once in our down-sampled dataset.

```{r}
# Down-sample down to just this % of our full data set
sample_proportion <- 0.2

# Sample size, round up.
sample_size <- ceiling(
  sample_proportion * nrow(starwars)
)
```

We can use `slice_sample` from the `dplyr` package to randomly sample star wars characters.

```{r}
naive_sampled_data <- starwars %>%
  slice_sample(
    n=sample_size,
    replace=FALSE
  )
naive_sampled_data
```

Take a look at `naive_sampled_data`.
What are some ways that our random sample might not be representative of our larger data set (`starwars`)?

### Exercises

Run the code below that calculates the frequencies of homeworlds in our _original_ (unsampled) dataset

```{r}
# Get the distribution of homeworlds in full dataset.
orig_homeworlds <- starwars %>%
  group_by(
    homeworld
  ) %>%
  summarise(
    n=n()
  ) %>%
  mutate(
    freq = n / sum(n),
  ) %>%
  arrange(
    desc(freq)
  )
orig_homeworlds
```

Run the code below to graph the frequencies of homeworlds in the original dataset.

```{r}
ggplot(
    orig_homeworlds,
    aes(
      x=reorder(homeworld, -freq),
      y=freq
    )
  ) +
  geom_bar(stat="identity") +
  labs(
    x="Homeworld",
    y="Frequency",
    title="Distribution of homeworlds for original dataset"
  ) +
  theme(
    axis.text.x=element_text(angle=45, hjust=1)
  )
```

- Calculate the frequencies of homeworlds in `naive_sampled_data`
- Graph the frequencies of homeworlds in `naive_sampled_data`
- What differences do you notice between the distribution of homeworlds in the original dataset and the sampled dataset?
- What happens to the distribution of homeworlds in the sample if you re-run all of the code above with different `sample_proportion` values?
- How could you modify `slice_sample` to randomly sample, but instead of weighting all rows equally, you weight each row's likelihood of being included in the sample according to the height attribute?
  Hint: `?slice_sample`

## Stratified random sampling

Stratified sampling can be used in cases where there are predefined groupings of objects.
In stratified sampling, you enforce the level of representation for each group.

For example, we might want to ensure that each species among Star Wars characters is represented in our sample.
There are many ways to implement this, and the particular approach you take will depend on your objective.
In this example, I'll randomly sample each species independently, guaranteeing that at least one character of each species is represented in our sample.

```{r}
# Grab the set of species represented in our original dataset
species <- levels(as.factor(starwars$species))
# Create an empty data frame that we will store our sample
stratified_sampled_data <- data.frame(
  matrix(
    ncol=ncol(starwars),
    nrow=0,
    dimnames=list(NULL,colnames(starwars))
  )
)
for (s in species) {
  # First, filter down to the species
  species_sample <- starwars %>%
    filter(
      species==s
    )
  # Calculate sample size, guarantee >= 1
  sample_size = max(
    ceiling(nrow(species_sample)*sample_proportion),
    1
  )
  # Sample from characters of the current species
  species_sample <- species_sample %>%
    slice_sample(
      n=sample_size,
      replace=FALSE
    )
  # Add sample of this species to larger sample
  stratified_sampled_data <- rbind(
    stratified_sampled_data,
    species_sample
  )
}
stratified_sampled_data
```

### Exercises

- Graph the distribution of species in the original `starwars` dataset and in the stratified sample (just like we did for homeworlds in the naive random sample). What differences do you notice? What might be causing those differences?
- We specified a sample proportion of `r sample_proportion`. Why might the size of our stratified sample differ from the size of our random sample?