---
output:
  pdf_document: default
  html_document: default
---

# In-class activity: visualizing data with ggplot

These activities are an opportunity to practice using ggplot to visualize data.

## Overview

### Goals

- Practice using ggplot to create the following kinds of data visualizations:
  - Boxplot
  - Line graph
  - Bar plot

### Sources

For more information/documentation on ggplot, I recommend these resources:

- [ggplot package documentation](https://ggplot2.tidyverse.org/reference/index.html)
- [ggplot cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)
- [Data visualization chapter in R for Data Science](https://r4ds.had.co.nz/data-visualisation.html)

## Setup

You should have already installed the `tidyverse` collection of packages, which includes the ggplot2 package.
If you haven't, go ahead and run:

```
install.packages("tidyverse")
```

You can now load all of the tidyverse packages, including ggplot2 with:

```{r}
library(tidyverse)
```

## Making a boxplot

A boxplot (sometimes called a box and whiskers plot) displays the distribution of a continous variable.
An individual boxplot divides a distribution into quartiles and visualizes them in a standardized way.
For more information about the "anatomy" of a boxplot, see: <https://clauswilke.com/dataviz/boxplots-violins.html#fig:boxplot-schematic>

With ggplot, you can use `geom_boxplot` to create a boxplot.

For the next few challenge problems, we'll make use of the `mpg` dataset.
Run `?mpg` for more information about the `mpg` dataset.

```{r}
mpg
```

**Challenge problem:** use ggplot to visualize highway fuel efficiency (`hwy`) as a function of vehicle class `class`.
I.e., your x-axis should be vehicle class, and your y-axis should be highway fuel efficiency.

```{r}
# Your code here
```

**Challenge problems:** make the following modifications to your boxplot

- Label the x-axis
- Label the y-axis
- Overlay a jitter plot (`geom_jitter`)
- Once you have a jittler plot layered over your boxplot, remove the outliers from being drawn in your boxplot (Hint: run and read through `?geom_boxplot`).
  - Why do we want to remove the outliers from being drawn on the geom_boxplot layer?
- Modify your plot to also map class to the color aesthetic
- Change the color palatte
- `facet_wrap` by manufacturer
  - Why are some of the facets missing boxes for some car classes?

## Making a line graph

Now, let's try making a line graph. For this we'll use the `storms` dataset is loaded when you load the tidyverse packages.
Read more about the `storms` dataset by running `?storms`.

```{r}
storms
```

We'll make a line graph showing the number of hurricanes by year.
So first, let's do some data wrangling on the `storms` dataset:

```{r}
# Filter down to just the storms categorized as hurricanes
hurricanes <- storms %>% filter(status == "hurricane")
# Use the group_by and summarize dplyr functions to count the number
# of hurricanes per year.
hurricane_cnt <- hurricanes %>%
  group_by(year) %>%
  summarise(
    year_n = n()
  )
```

**Challenge problem:** Use `geom_line` to create a line graph of hurricanes over time.

```{r}
# Your code here
```

**Challenge problems:** Modify your line graph as follows

- Label your x and y axis
- Add a title
- Change the solid line to a dashed line

## Making a bar plot

Next, we'll make a bar plot. For this, we'll try plotting the distribution of homeworlds of the human characters in the `starwars` dataset.
The `starwars` dataset is loaded into R when you load the `tidyverse` collection of packages.

```{r}
starwars
```

First, let's do some data wrangling on the starwars dataset to
(1) filter down to just human characters and
(2) count how many characters are from each homeworld.

```{r}
sw_humans <- starwars %>% filter(species == "Human")
sw_homeworlds <- sw_humans %>%
  group_by(homeworld) %>%
  summarize(
    n = n()
  )
```

```{r}
sw_homeworlds
```

**Challenge problem:** use `geom_col` to plot the number of human characters from each planet (use the `sw_homeworlds` dataframe).

```{r}
# Your code here
```

**Challenge question:** What does the NA column represent?

**Challenge problem:** use `geom_bar` to plot the number of human characters from each planet (use the `sw_humans` dataset)

```{r}
# Your code here
```

**Challenge question:** Explain the difference between `geom_bar` and `geom_col`

