---
output:
  pdf_document: default
  html_document: default
---

# In-class activity: using tidyr to create tidy data

## Overview

As part of this activity, you will download three "messy" data sets and use tidyverse functions to transform them into "tidy" data.
If you want to read more about tidy data, I suggest the following article for an introduction to tidy data: <https://tidyr.tidyverse.org/articles/tidy-data.html>.

The tidyr package provides functions for making messy data "tidy".
For a complete list of the included functions, see <https://tidyr.tidyverse.org/reference/index.html>.

Understanding tidy data and how to reformat your data into a tidy format is very important, as many of the functions/packages that we will use this semester are written to work with tidy-format data.
For example, ggplot (the R graphics package that we will be using) assumes tidy data.
Taking the time now to get comfortable with what it means to have tidy data may save you substantial time later on.

### Goals:

- Work through examples of using tidyr (and dplyr) functions to clean up messy (un-tidy) data

### Sources

- <https://tidyr.tidyverse.org/articles/tidy-data.html>
- [tidyr functions documentation](https://tidyr.tidyverse.org/reference/index.html)
- [R for Data Science](https://r4ds.had.co.nz/tidy-data.html)

## Setup

You should have already installed the `tidyverse` collection of packages, which includes the tidyr package.
But, if you haven't, go ahead and run:

```
install.packages("tidyverse")
```

We can load all of the tidyverse packages into R using:

```{r}
library(tidyverse)
```

## Messy grade data

### What is tidy data?

You can represent the same underlying data many ways.
Download the two messy grade data sets from blackboard.

**Challenge problem:** Load messy_grades_1.csv into R:

```{r}
# Your code here
```

**Challenge problem:** Load messy_grades_2.csv into R:

```{r}
# Your code here
```

**Challenge questions:**

- Why are these data sets (messy_grades_1 and messy_grades_2) not tidy?
- Sketch out a new way to format the grade data that is tidy.
- Which tidyr function can you use to reformat the messy grade data? (look through the tidyr functions documentation linked above)

### Making data tidy

**Challenge problem:** Use the function you identified previously to create reformat the grade data into a tidy format:

```{r}
# Your code here
```

## Messy temperature data

**Challenge problem:** Load the messy temperatures data into R:

```{r}
# Your code here
```

**Challenge questions:**

- Why is this temperature data set messy?
- Sketch out a new way to format the temperature data so that it is tidy.
- Which tidyr function can you use to help you reformat the messy temperature data?

**Challenge problem**: Use the function you identified to reformat the temperature data into a tidy format.

```{r}
# Your code here
```

