---
output:
  pdf_document: default
  html_document: default
---

# In-class activity: using dplyr to manipulate data

These activities are an opportunity to practice using dplyr functions to wrangle data.

## Goals:

- Learn about solving common data manipulation challenges with dplyr
- Learn to use the pipe operator (`%>%`)

## Sources:

For more information/documentation on data wrangling with dplyr, I recommend these resources:

- [Introduction to dplyr](https://dplyr.tidyverse.org/articles/dplyr.html)
- [R for Data Science](https://r4ds.had.co.nz/transform.html)
- [dplyr documentation](https://dplyr.tidyverse.org/)
- [Software carpentry lesson on dplyr](https://swcarpentry.github.io/r-novice-gapminder/13-dplyr/index.html)

### dplyr cheat sheet

Download pdf here: <https://github.com/rstudio/cheatsheets/blob/main/data-transformation.pdf>

## Setup

You might have already installed the `tidyverse` collection of packages, which includes the dplyr package.
But, if you haven't, go ahead and run:

```
install.packages("tidyverse")
```

With the tidyverse packages installed, you'll need to load the dplyr library to use it.
You can load the dplyr package individually using `library(dplyr)`.
Alternatively, you can load all tidyverse packages:

```{r}
library(tidyverse)
```

## Overview

This document will cover key dplyr functions that solve the most common data manipulation challenges.
The examples and explanations here non-exhaustive (dplyr has a lot of functionality and each of these functions are very flexible).
Instead, I'd like you to start building a sense of the broad category of data manipulations that are possible with each of the dplyr functions below.
As you encounter challenges with manipulating data, you should have a solid idea of which functions you will need to do what you want to do.
Then, you can refer to the appropriate documentation to work out the specific details.

dplyr's functions for basic data manipulation can be divided into three categories based on the component of the dataset that they work with (descriptions pulled from [dplyr documentation](https://dplyr.tidyverse.org/articles/dplyr.html)):

- Rows
  - `filter()` - choose rows in your data based on their column values
  - `slice()` - choose rows in your data based on their location (index) in the dataset
  - `arrange()` - change the ordering of rows in your data
- Columns
  - `select()` - changes whether or not a column is included
  - `rename()` - changes the names of columns
  - `mutate()` - changes the values of columns and can create new columns that are functions of existing columns
  - `relocate()` - changes the order of columns
- Groups of rows
  - `summarise()` - collapses a group of rows into a single row

These functions can be used in combination with `group_by()`, which allows you to perform these functions on particular groups of rows, instead of the entire dataset.

Note that the above list of functions is non-exhaustive.
I refer you to the [dplyr documentation](https://dplyr.tidyverse.org/reference/index.html) for an exhaustive list.
Additionally, I'd encourage you to look over (and save) the cheat sheet linked at the end of this document.

## The pipe operator

Most dplyr functions take a dataframe (or tibble) as their first argument.
dplyr provides the pipe operator (`%>%`), which lets you conveniently compose multiple dplyr functions in a single line (instead of needing to save intermediate states of your dataset).
The pipe operator "pipes" the results of the left side of the pipe into the first argument of the right side of the pipe.

```{r}
f <- function(a, b) {
  return(a + b)
}
g <- function(a, b) {
  return(a * b)
}

x <- 2
y <- 100
```

**Challenge problem:** In the code chunk below, use the pipe operator to call the `f` with `x` and `y` as arguments.

```{r}
# Your code here
```

**Challenge problem:** In the code chunk below, chain together two pipe operators to call `g(f(x, y), y)`

```{r}
# Your code here
```

## Data

We'll use the `starwars` dataset to demonstrate the basic data manipulation functions provided by dplyr.
You don't need to download anything, the `starwars` dataset is loaded when you load the dplyr package.

```{r}
dim(starwars)
```

This dataset has 87 rows and 14 columns where each row represents a character from Star Wars, and each column is a particular attribute.

```{r}
starwars
```

If you look closely at the above output, you'll notice that the `starwars` dataset is represented as a tibble.
A tibble is a more modern version of the dataframe type, and for the most part, anywhere you could use a dataframe, you can use a tibble.
You can convert a data frame to tibbles with the `as_tible()` function.
Read more about tibbles here: <https://tibble.tidyverse.org/>.

## Filter rows with `filter()`

`filter()` allows you to choose rows in a data frame based on their column values.

For example, if we wanted to create a new data frame that only contained droids, we could:

```{r}
starwars %>% filter(species=="Droid")
```

We can also filter rows based on more complex conditions:

```{r}
starwars %>% filter(species == "Droid" & height > 100)
```

**Challenge problem:** Use the filter function to filter the data to only characters with the following species attribute: Droid, Wookiee, and Hutt.

```{r}
# Your code here
```

Notice that the original `starwars` dataset is never modified by any of the filter functions.
None of the dplyr functions modify the dataset passed to them with the `data` argument.
Instead, they return a new data frame with the appropriate manipulations.

## Arrange rows with `arrange()`

`arrange()` reorders rows in the given data frame.
It takes a data frame and a set of columns names to reorder by.
If you give more than one column name, additional column names are used to break ties in the values of proceeding column names.

**Challenge problem:** Use the arrange function to order the characters by their mass. Which character has the greatest mass?

```{r}
# You code here
```

## Choose rows using their position with `slice()`

Use slice to choose rows by their position.

For example, to choose the character in the second row of the data frame:

```{r}
starwars %>% slice(2)
```

The first, tenth, and twentieth rows of the data frame:

```{r}
starwars %>% slice(c(1,10,20))
```

The odd rows of the data frame:

```{r}
starwars %>% slice(seq(1,nrow(starwars), 2))
```

**Challenge problem:** Use the slice and arrange functions to quickly get the character with the largest height.

```{r}
# Your code here
```

There are a few convenient helper versions of the slice function:

- `slice_head`
- `slice_tail`
- `slice_sample`
- `slice_min`
- `slice_max`

**Challenge problem:** Use the `slice_max` function to quickly get the character with the largest mass.

```{r}
# Your code here
```

## Select columns with `select()`

When you're working with large datasets, sometimes you want to drop all of the columns you're not using or interested in.

For example, if you only wanted the species column:
```{r}
starwars %>% select(species)
```

Or, if you wanted name, species, and homeworld:

```{r}
starwars %>% select(name, species, homeworld)
```

What if you wanted all columns with the letter m in the column name?

```{r}
starwars %>% select(contains("m"))
```

See [the tidyselect documentation](https://tidyselect.r-lib.org/reference/starts_with.html) for more selection helpers (like contains).

**Challenge problem:** use the select function to select all columns that end in "s"

```{r}
# Your code here
```

## Rename columns with `rename()`

`rename()` lets you rename columns.

**Challenge problem:** Use the rename function to rename the column "homeworld" to "party_planet":

```{r}
# Your code here
```

## Add new columns with `mutate()`

Often, it is useful to add new columns that are functions of existing columns.

**Challenge problem:** Use the mutate function to add a new column, "heightXmass".
The value in this column for each row should be the product of that row's height and mass values.

```{r}
# Your code here
```

```{r}
colnames(starwars)
```

## Summarize groups of rows using `summarise()`

`summarise()` collapses a dataset into a single row.

```{r}
starwars %>% summarise(mass = mean(mass, na.rm=TRUE))
```

`summarise()` is particularly useful in combination with the `group_by` function.

For example, to get the average mass of characters in each species,
```{r}
starwars %>% group_by(species) %>% summarise(mass=mean(mass, na.rm=TRUE))
```

Or, if we wanted to count the number of characters of each species,

```{r}
starwars %>% group_by(species) %>% summarise(count=n())
```




