---
output:
  pdf_document: default
  html_document: default
---

# Homework 3

Names: Rocket Raccoon, Groot

## Part A. Use R as a calculator

### Question 1

```{r}
# Your code here
```

### Question 2

```{r}
# Your code here
```

### Question 3

```{r}
# Your code here
```

## Part B. Built-in Functions

## Question 1

```{r}
# Your code here
```

## Question 2

```{r}
# Your code here
```

## Question 3

```{r}
# Your code here
```

## Question 4

```{r}
# Your code here
```


## Part C. Creating vectors

## Question 1

```{r}
# Your code here
```

## Question 2

```{r}
# Your code here
```

## Question 3

```{r}
# Your code here
```


## Part D. Subsetting vectors

## Question 1

```{r}
# Your code here
```

## Question 2

```{r}
# Your code here
```

## Question 3

```{r}
# Your code here
```


## Part E. Types of data

## Question 1

```{r}
# Your code here
```

## Question 2

```{r}
# Your code here
```

## Question 3

```{r}
# Your code here
```

## Question 4

```{r}
# Your code here
```

## Question 5

```{r}
# Your code here
```

## Question 6

```{r}
# Your code here
```

## Question 7

```{r}
# Your code here
```

## Part F. Control flow

## Question 1

```{r}
# Your code here
```

## Part G. Tidyverse

## Question 1


## Question 2


## Question 3


